### 2016-05-11 text/Chapter02.tex
Add binding to chapter 2, whitespace, tweaks

@@

We start by looking at coeffects as the dual of effect systems (Section~\ref{sec:path-eff}). Next,
we extend the duality to category theory, looking at the categorical dual of monads known as 
\emph{comonads} (Section~\ref{sec:path-sem}). Finally we look at logically inspired type systems 
that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).

%===================================================================================================

@@

\subsection{Simple coeffect system.}
\label{sec:path-effects-coeff}
When writing the judgements of coeffect systems, we want to emphasize the fact that coeffect 
systems talk about \emph{context} rather than \emph{results}. For this reason, we write the 
judgements in the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$, associating the additional 
information with the context (left-hand side) of the judgement rather than with the result 
(right-hand side) as in $\Gamma \vdash e : \tau, \cclrd{r}$. This change alone would not be 
very interesting -- we simply used different syntax to write a predicate with four arguments. 
The more interesting difference is how the lambda abstraction rule looks.

The language in Figure~\ref{fig:path-coeff} extends simple lambda calculus with resources and
with a construct $\kvd{access}~e$ that obtains the resource specified by the expression $e$.
Most of the typing rules correspond to those of effect systems. Variable access (\emph{var}) 
has no context requirements, application (\emph{app}) combines context requirements of the two
sub-expressions and latent context-requirements of the function.

However, the (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
resource requirements of the body $\cclrd{r} \cup \cclrd{s}$ are split between the \emph{immediate 
context-requirements} associated with the current context $\coctx{\Gamma}{\cclrd{r}}$ and the 
\emph{latent context-requirements} of the function.

As will be discussed in more detail in Section~\ref{sec:applications-flat-impl}, this means that a 
resource can be captured when a function is declared (e.g.~when it is constructed on the 
server-side where database access is available), or when a function is called (\eg~when a function 
created on server-side requires access to current time-zone, it can use the resource available on
 the client-side). In other words, resources in this example support both static (lexical) and
 dynamic scoping.

@@

%===================================================================================================

@@

\noindent
Without providing much details, we note that well known examples of monads include the partiality
monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in 
Haskell, list monad ($\mtyp{}{\tau} = \mu \gamma.1 + (\tau \times \gamma)$) and other.
In programming language semantics, monads can be used in two distinct ways.

@@

In this thesis, we generally follow the first approach -- this means that we work with an existing
programming language without needing to add additional constructs corresponding to the primitives
of our semantics (the alternative is discussed in Section~\ref{sec:unified-meta}). To clarify the 
difference, the following two sections show a minimal example of both formal systems. We follow 
Moggi and start with language where judgements have the form $x\!:\!\tau_1 \vdash e : \tau_2$ with 
exactly one variable\footnote{This simplifies the examples as we do not need \emph{strong} monad, 
but that is an orthogonal issue to the distinction between language semantics and meta-language.}.

@@

\paragraph{Language semantics.} When using monads to provide semantics of a language, we do not
need to extend the language in any way -- we assume that the language already contains the 
effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
of the form $x\!:\!\tau_1 \vdash e : \tau_2$ is interpreted as a morphism $\tau_1 \rightarrow \mtyp{}{\tau_2}$,
meaning that any expression is interpreted as an effectful computation. The semantics of variable
access and the application of a primitive function $f$ is interpreted as follows:
%
\begin{equation*}
\begin{array}{rcl}
 \sem{x\!:\!\tau_1 \vdash x : \tau_1} &=& \ident{unit}_\mtyp{}{}\\
 \sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
\end{array}
\end{equation*}
%
Variable access is an effect-free computation, that returns the value of the variable, wrapped
using $\ident{unit}_\mtyp{}{}$. In the second rule, we assume that $e$ is an expression using
the variable $x$ and producing a value of type $\tau_2$ and that $f$ is a (primitive) function
$\tau_2 \rightarrow \mtyp{}{\tau_3}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
to a function $\mtyp{}{\tau_2} \rightarrow \mtyp{}{\tau_3}$ which is compatible with the 
interpretation of the expression $e$.

@@

\paragraph{Meta-language interpretation.} ~When designing meta-language based on monads, we need to
extend the lambda calculus with additional type(s) and expressions that correspond to monadic
primitives. Assuming $\alpha$ is a primitive type:
%
\begin{align*}
\tau &:= \alpha \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
   e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
\end{align*}
%
The types consist of the primitive type, function type and a type constructor that 
represents monadic computations. This means that the expressions in the language can create both
effect-free values, such as $\tau$ and computations $\mtyp{}{\tau}$. The additional expression
$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no actual effects) from a
value and $\kvd{let}_\mtyp{}{}$ is used to sequence effectful computations. In the semantics, 
monads are not needed to interpret variable access and application, they are only used in the 
semantics of additional (monadic) constructs:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \idf{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\tau_2}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\tau_3}} &=& 
  \ident{bind}_\mtyp{}{}~\sem{e_2} \circ \sem{e_1}
\end{array}
\end{equation*}

@@

Moggi regards the meta-language system as more fundamental, because \emph{``its models are more 
general''}. Indeed, this is a valid and reasonable perspective. Yet, we follow the first style,
precisely because it is \emph{less general}. As discussed in Chapter~\ref{ch:intro}, our aim is 
to develop concrete context-aware programming languages (together with their type theory and 
semantics) rather than to build a general framework for reasoning about languages with 
context-dependent properties.

@@

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$ 
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta) 
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:
\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

@@

\paragraph{Language semantics.}
To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
$\ctyp{}{\tau} = \mu \gamma.\tau + (\tau \times \gamma)$. A value of the type is either
the last element $\tau$ or an element followed by another non-empty list $\tau \times \gamma$
(consisting of the head $\tau$ and the tail $\gamma$). Note that the list must be non-empty, 
otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In 
the following, we write $(l_1, \ldots, l_n)$ for a list of $n$ elements:
%
\begin{equation*}
\begin{array}{rcl}
\ident{counit}~(l_1, \ldots, l_n) &=& l_1\\
\ident{cobind}~f~(l_1, \ldots, l_n) &=& (f (l_1, \ldots, l_n), f (l_2, \ldots, l_n), \ldots, f (l_n))
\end{array}
\end{equation*}
%
The \ident{counit} operation returns the current (first) element of the (non-empty) list.
The \ident{cobind} operation creates a new list by applying the context-dependent function $f$
to the entire list, to the suffix of the list, to the suffix of the suffix and so on. Interestingly,
it preserves the \emph{shape} of the list as it turns a list of $n$ elements into another list
of $n$ elements.

@@

Now, consider a simple data-flow language with single-variable contexts, variables, 
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$ 
has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
the language models data-flow and values are lists (of past values) is a matter of semantics,
which is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (obtain the value and ignore additional available context); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).

For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an 
expression that expects context with variable $x$ and returns a stream of values before the 
previous one. The semantics of the term expresses this behaviour: 
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed 
earlier.

@@

We return to this topic when discussing lambda abstraction in Section~\ref{sec:applications-structure-lam} 
and semantics of flat coeffect systems in Section~\ref{sec:flat-semantics}.

@@

\paragraph{Meta-language interpretation.} ~To briefly demonstrate the app\-roach that employs comonads
as part of a meta-language, we look at an example inspired by the work of Pfenning, Davies and 
Nanevski. We do not attempt to provide a precise overview of their work. The main purpose 
of the following discussion is to provide a different intuition behind comonads, and to present an 
example of a language that includes comonad as a type constructor, together with language primitives
corresponding to comonadic operations\footnote{In fact, Pfenning, Davies and Nanevski 
\cite{logic-modal-reconstruction,logic-cmtt} never mention comonads explicitly. This is done in later 
work by Gabbay et al. \cite{logic-cmtt-semantics},  but the connection between the language and comonads
is not as direct as in case of monadic or comonadic semantics covered in the previous section.}.

@@

In the coeffect system for tracking resource usage outlined earlier, we associated additional
contextual information (set of available resources) with the variable context of the typing 
judgement: $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$. In other words, our work focuses on 
``what is happening on the left hand side of $\vdash$''.

>>>>

We start (Section~\ref{sec:path-binding}) by discussing how coeffects arise from the most common
notion of context-dependence -- variable binding. Next, we look at coeffects as the dual of effect
systems (Section~\ref{sec:path-eff}) and we extend the duality to category theory, looking at
\emph{comonads} (Section~\ref{sec:path-sem}). Finally we consider logically inspired type systems
that are closely related to our structural coeffects (Section~\ref{sec:path-logic}).

%===================================================================================================
%
%    ######
%    #     # # #    # #####  # #    #  ####
%    #     # # ##   # #    # # ##   # #    #
%    ######  # # #  # #    # # # #  # #
%    #     # # #  # # #    # # #  # # #  ###
%    #     # # #   ## #    # # #   ## #    #
%    ######  # #    # #####  # #    #  ####
%
%===================================================================================================

\section{Through static and dynamic binding}
\label{sec:path-binding}

Accessing a variable is arguably the simplest form of context-dependence, to the extent that we
do not normally think of variables as a notion of context. However, variables fit well with our
earlier description of context in programming: a block of code that accesses a variable can only
be executed in an environment where the variable is available.

In this section, we look at variable binding through the perspective of context-requirements.
We discuss ordinary variable binding and Haskell's implicit parameters \cite{app-implicit-parameters},
which provides an interesting point in the design space. We then discuss ambiguity that is inherent
in variable binding and, more generally, context-aware programming languages, and a way of
resolving it through \emph{type-directed semantics}.

@@

\subsection{Variable binding}
\label{sec:path-binding-var}

Variable access represents a form of context-dependence. For example, an expression $x+y$ can be
only evaluated if the environment provides values for variables $x$ and $y$. A variable
\emph{requirements} can be satisfied in two standard ways that are captured by \emph{dynamic} and
\emph{static} (or lexical) variable binding. Consider the following simple program:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The program can be evaluated in two ways, depending on the variable binding mechanism:
%
\begin{itemize}
\item \textsc{Static (lexical) binding.} In a langauge with static binding (such as ML or Java),
  the variable \ident{x} inside the body of the lambda function is statically bound to the
  declaration in the lexical scope -- that is, the variable on the second line -- and the
  expression evaluates to 10.

\item \textsc{Dynamic binding.} In a language with dynamic binding (some variants of LISP), the
  variable value is dynamically bound to the last value that has been added to the environment
  during program execution. Thus, the \ident{x} variable inside the lambda function refers to
  \ident{x} defined on line 5 of the sample and the expression evaluates to 5.
\end{itemize}

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context requirements of the body can be placed on the scope in
which the function is defined (declaration-site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call-site).

In static and dynamic scoping, all variable requirements are always placed on one site. However,
those are not the only two options. It is conceivable that a language would use a mechanism that
splits variable requirements differently and combines aspects of dynamic and static binding. For
example, the language could use statis binding by default, but resort to dynamic binding if a
variable is not available in the lexical scope. One such system is discussed in the next section.

Languages with static scoping generally check that all variable requirements are satisfied.
Attempt to access a variable that is not available in scope will give a compile error. In contrast,
languages with dynamic variable binding typically do not perform compile-time checks -- if a
program attempts to access a variable that is not available in the environment, a runtime error
occurs. This is possibly artifact of their implementation -- implementing static binding
\emph{without} checking would be cumbersome and implementing checking for dynamic binding is
requires a more sophisticated type system (Section~\ref{sec:path-effects-coeff}).

\subsection{Implicit parameter binding}

Haskell uses static binding for ordinary variables, but it additionally provides a feature named
\emph{implicit parameters} \cite{app-implicit-parameters} that adds a special kind of variables,
written as \ident{?param}, which use a particular combination of static and dynamic binding.

The following two examples are variations on the one discussed in the Section~\ref{sec:path-binding-var}.
We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
implicit parameter is defined both at declaration-site and at the call-site. On the right, the
implicit parameter is available only at the call-site:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call-site.
In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
delayed context requirements and the expression evalautes to $10$. In the example on the right,
the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

In Haskell, the type system checks that bindings for all required implicit parameters are
available. The type of the function \ident{f} on the left is $\kvd{int}\rightarrow\kvd{int}$, while
the type of the \ident{f} function on the right is $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
in the environment when calling the function. It is worth noting that the syntax is similar to the
one used by type-class constraints. Those can be viewed as context requirements too
(Section~\ref{sec:applications-flat-impl}).

\paragraph{Thesis perspective.}
The three different binding mechanisms discussed so far can be seen as different ways of splitting
context requirements of a particular kind. Dynamic and static binding represent the opposite ends
of the design space and Haskell's implicit parameters are an interesting point inside the wider space.

In this thesis, we consider different notions of context, using implicit parameters as just one
of several motivating examples. However, implicit parameters are a valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context requirements of
a function can be satisfied using the context available at declaration-site or using the context
available at the call-site.

We also aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context requirements are split between the declaration-site and the call-site. While
Haskell's solution for implicit parameters might be the most reasonable one, different notions
of context might require different domain-specific choices and the general framework of
context-aware programming should make that possible.

\subsection{Resolving ambiguity}
\label{sec:path-binding-amb}

In many practical programming languages, the value and semantics of an expression depends on the
type derivation. Typically, the choice is hidden behind a mechanism that selects one preferred type
derivation.

This mechanism serves as an inspiration for our approach to resolving ambiguity inherent in
context-aware programs. This section shows a brief example using the F\# language \cite{app-fsharp},
before revisiting the implicit parameters example.

~

Consider an F\# lambda expression $(\lambda\ident{x}\rightarrow\ident{x.Length})$, which takes
an object $\ident{x}$ and returns the value of its \ident{Length} property. F\# is a nominally-typed
language meaning that, in isolation, the function has an ambiguous meaning\footnote{In contrast,
in a structurally-typed language, the function would have a unique typing in isolation. In OCaml,
the type would be $\langle \ident{Length}:\ident{'a}\rangle \rightarrow \ident{'a}$.}. It can be
a function taking an array, it can be a function taking a string, or it can be a function taking
one of the other .NET types that are equipped with the \ident{Length} property.

The semantics of the function depends on the typing derivation. For example, for arrays, it is
compiled using the \ident{ldlen} intermediate language (IL) instruction, while or strings, it is
compiled using \ident{call} instruction (calling the property getter). In F\#, the compiler chooses
an appropriate typing derivation. For example:
%
\begin{equation*}
\begin{array}{l}
  \lbrack \str{hello};\str{world} \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
  \lbrack \ident{Array.empty};~\ident{Array.create}~\num{100}~\num{0}\; \rbrack ~|\hspace{-0.25em}>\;\ident{List.map}~(\lambda\ident{s}\rightarrow\ident{s.Length})\\
\end{array}
\end{equation*}
%
The $|\hspace{-0.25em}>$ operator passes the value on the left to the function on the right. In
the first case, the compiler infers that the type of the input is a list of strings and so the
type of the lambda function becomes $\ident{string}\rightarrow\ident{int}$. In the second case,
the list contains two arrays (empty array and array containing one hundred 0 values) and so
the type of the lambda function is $\ident{int}\lbrack\,\rbrack\rightarrow\ident{int}$.
The important points about the example are:
%
\begin{itemize}
\item The semantics of the function $(\lambda\ident{x}\rightarrow\ident{x.Length})$
  depends on its type. For arrays, it is compiled using a special IL instruction, while for
  strings, it calls a property getter.

\item The compiler chooses an appropriate typing derivation. In the above case, this is done
  based on the context in which the expression appears, but other options are possible
  (in some cases, there is a \emph{default} resolution; in some cases, the compiler requires
  explicit typing annotation).
\end{itemize}
%
The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
possible typing derivations and its semantics varies depending on the type. If the lexical
scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
and it captures the value from the lexical scope. Otherwise, the type of the function is
$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
value from a hidden dictionary that is passed together with the input from the call-site.

\paragraph{Thesis perspective.}
Just like the F\# function in the above example, certain expressions in context-aware languages
developed in this thesis have multiple valid typing derivations and their semantics depends on the
type. In F\#, the compiler determines a unique typing derivation based on other parts of the program
(or fails, if type is not uniquely determined). In our languages, we also determine a unique typing
derivation. However, rather than relying on type information from other parts of the program, we
explicitly define algorithm that chooses the preferred unique derivation.

This approach decouples two important aspects of context-aware programming and lets us study them
independently -- the semantics of context-aware programs and the domain-specific way of resolving
how context requirements are satisfied. In our treatment of implicit parameters, we consider
multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
behaviour of Haskell).

%===================================================================================================
%
%    #######
%    #       ###### ###### ######  ####  #####     ####  #   #  ####  ##### ###### #    #  ####
%    #       #      #      #      #    #   #      #       # #  #        #   #      ##  ## #
%    #####   #####  #####  #####  #        #       ####    #    ####    #   #####  # ## #  ####
%    #       #      #      #      #        #           #   #        #   #   #      #    #      #
%    #       #      #      #      #    #   #      #    #   #   #    #   #   #      #    # #    #
%    ####### #      #      ######  ####    #       ####    #    ####    #   ###### #    #  ####
%
%===================================================================================================

@@

\subsection{Simple coeffect system.}
\label{sec:path-effects-coeff}

When writing the judgements of coeffect systems, we want to emphasize the fact that coeffect
systems talk about \emph{context} rather than \emph{results}. For this reason, we write the
judgements in the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$, associating the additional
information with the context (left-hand side) of the judgement rather than with the result
(right-hand side) as in $\Gamma \vdash e : \tau, \cclrd{r}$. This change alone would not be
very interesting -- we simply used different syntax to write a predicate with four arguments.
The more interesting difference is how the lambda abstraction rule looks.

The language in Figure~\ref{fig:path-coeff} extends simple lambda calculus with resources and
with a construct $\kvd{access}~e$ that obtains the resource specified by the expression $e$.
Most of the typing rules correspond to those of effect systems. Variable access (\emph{var})
has no context requirements, application (\emph{app}) combines context requirements of the two
sub-expressions and latent context-requirements of the function.
The (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
resource requirements of the body $\cclrd{r} \cup \cclrd{s}$ are split between the \emph{immediate
context-requirements} associated with the current context $\coctx{\Gamma}{\cclrd{r}}$ and the
\emph{latent context-requirements} of the function.

This is where context-aware languages permit multiple valid typing derivations as discussed in
Section~\ref{sec:path-binding-amb}. In the example here, a resource can be captured
when a function is declared (e.g.~when it is constructed on the server-side where database access
is available), or when a function is called (\eg~when a function created on server-side requires
access to current time-zone, it can use the resource available on the client-side). In other words,
resources in this example support both static (lexical) and dynamic scoping. Out of the multiple
valid typing derivation, we would choose one -- for example, capturing only those server-side
resources that are not available on the client-side\footnote{This can be characterized as
``\emph{dynamic binding when possible, static binding when needed}'' and it is, quite curiously, the
opposite choice than the one used by Haskell's implicit parameters.}. We discuss this
system in detail in Section~\ref{sec:applications-flat-impl}.

@@

%===================================================================================================
%
%     #####
%    #     # ###### #    #   ##   #    # ##### #  ####   ####
%    #       #      ##  ##  #  #  ##   #   #   # #    # #
%     #####  #####  # ## # #    # # #  #   #   # #       ####
%          # #      #    # ###### #  # #   #   # #           #
%    #     # #      #    # #    # #   ##   #   # #    # #    #
%     #####  ###### #    # #    # #    #   #   #  ####   ####
%
%===================================================================================================

@@

\noindent
Without providing much details, we note that well known examples of monads include the partiality
monad ($\mtyp{}{\alpha} = \alpha + {\bot}$) also corresponding to the \ident{Maybe} type in
Haskell, list monad ($\mtyp{}{\tau} = 1 + (\tau \times \mtyp{}{\tau})$) and other.
In programming language semantics, monads can be used in two distinct ways.

@@

In this thesis, we generally follow the first approach -- this means that we work with an existing
programming language without needing to add additional constructs corresponding to the primitives
of our semantics (the alternative is discussed in Section~\ref{sec:further-meta}). To clarify the
difference, the following two sections show a minimal example of both formal systems. We follow
Moggi and start with language where judgements have the form $x\!:\!\tau_1 \vdash e : \tau_2$ with
exactly one variable\footnote{This simplifies the examples as we do not need \emph{strong} monad,
but that is an orthogonal issue to the distinction between language semantics and meta-language.}.

@@

\paragraph{Language semantics.} When using monads to provide semantics of a language, we do not
need to extend the language in any way -- we assume that the language already contains the
effectful primitives (such as the assignment operator $x \leftarrow e$ or other). A judgement
of the form $x\!:\!\tau_1 \vdash e : \tau_2$ is interpreted as a morphism $\tau_1 \rightarrow \mtyp{}{\tau_2}$,
meaning that any expression is interpreted as an effectful computation. The semantics of variable
access and the application of a primitive function $f$ is interpreted as follows:

\begin{equation*}
\begin{array}{rcl}
 \sem{x\!:\!\tau_1 \vdash x : \tau_1} &=& \ident{unit}_\mtyp{}{}\\
 \sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& (\ident{bind}_\mtyp{}{}~f) \circ \sem{e}\\
\end{array}
\end{equation*}
%
Variable access is an effect-free computation, that returns the value of the variable, wrapped
using $\ident{unit}_\mtyp{}{}$. In the second rule, we assume that $e$ is an expression using
the variable $x$ and producing a value of type $\tau_2$ and that $f$ is a (primitive) function
$\tau_2 \rightarrow \mtyp{}{\tau_3}$. The semantics lifts the function $f$ using $\ident{bind}_\mtyp{}{}$
to a function $\mtyp{}{\tau_2} \rightarrow \mtyp{}{\tau_3}$ which is compatible with the
interpretation of the expression $e$.

@@

\paragraph{Meta-language interpretation.} ~When designing meta-language based on monads, we need to
extend the lambda calculus with additional type(s) and expressions that correspond to monadic
primitives:
%
\begin{align*}
\tau &:= \ident{num} \sep \tau_1 \rightarrow \tau_2 \sep \mtyp{}{\tau} \\
   e &:= x \sep f~e \sep \kvd{return}_\mtyp{}{}~e \sep \kvd{let}_\mtyp{}{}~x \Leftarrow e_1~\kvd{in}~e_2
\end{align*}
%
The types consist of the primitive type, function type and a type constructor that
represents monadic computations. This means that the expressions in the language can create both
effect-free values, such as $\tau$ and computations $\mtyp{}{\tau}$. The additional expression
$\kvd{return}_\mtyp{}{}$ is used to create a monadic computation (with no actual effects) from a
value and $\kvd{let}_\mtyp{}{}$ is used to sequence effectful computations. In the semantics,
monads are not needed to interpret variable access and application, they are only used in the
semantics of additional (monadic) constructs:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \idf{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{return}_\mtyp{}{}~e : \mtyp{}{\tau_2}} &=& \ident{unit}_\mtyp{}{} \circ \sem{e}\\
\sem{x\!:\!\tau_1 \vdash \kvd{let}_\mtyp{}{}~y \Leftarrow e_1~\kvd{in}~e_2 : \mtyp{}{\tau_3}} &=&
  \ident{bind}_\mtyp{}{}~\sem{e_2} \circ \sem{e_1}
\end{array}
\end{equation*}

@@

Moggi regards the meta-language system as more fundamental, because \emph{``its models are more
general''}. This is a valid and reasonable perspective. Yet, we follow the first style,
precisely because it is \emph{less general}. Our aim is to develop concrete context-aware
programming languages (together with their type systems and semantics) rather than to build a
general framework for reasoning about languages with contextual properties.

@@

\begin{definition}
A \emph{comonad} over a category $\catc$ is a triple $(C, \ident{counit}, \ident{cobind})$ where:
\begin{compactitem}
\item $C$ is a mapping on objects (types) $C : \catc \rightarrow \catc$
\item $\ident{counit}$ is a mapping $\ctyp{}{\alpha} \rightarrow \alpha$
\item $\ident{cobind}$ is a mapping $(\ctyp{}{\alpha} \rightarrow \beta)
  \rightarrow (\ctyp{}{\alpha} \rightarrow \ctyp{}{\beta})$
\end{compactitem}
%
such that, for all $f:\ctyp{}{\alpha} \rightarrow \beta$ and $g:\ctyp{}{\beta} \rightarrow \gamma$:

\begin{align}
\tag{\emph{left identity}}
  \ident{cobind}~\ident{counit} &= \idf{}
  \\
\tag{\emph{right identity}}
  \ident{counit} \circ \ident{cobind}~f &= f
  \\
\tag{\emph{associativity}}
  \ident{cobind}~(g \circ \ident{cobind}~f) &= (\ident{cobind}~g) \circ (\ident{cobind}~f)
\end{align}
\end{definition}

@@

\paragraph{Language semantics.}
To demonstrate the approach of Uustalu and Vene, we consider the non-empty list comonad
$\ctyp{}{\tau} = \tau + (\tau \times \ctyp{}{\tau})$. A value of the type is either
the last element $\tau$ or an element followed by another non-empty list $\tau \times \ctyp{}{\tau}$
(consisting of the head $\tau$ and the tail $\ctyp{}{\tau}$). Note that the list must be non-empty,
otherwise \ident{counit} would not be a complete function (it would be undefined on empty list). In
the following, we write $(l_1, \ldots, l_n)$ for a list of $n$ elements:
%
\begin{equation*}
\begin{array}{rcl}
\ident{counit}~(l_1, \ldots, l_n) &=& l_1\\
\ident{cobind}~f~(l_1, \ldots, l_n) &=& (f (l_1, \ldots, l_n), f (l_2, \ldots, l_n), \ldots, f (l_n))
\end{array}
\end{equation*}
%
The \ident{counit} operation returns the current (first) element of the (non-empty) list.
The \ident{cobind} operation creates a new list by applying the context-dependent function $f$
to the entire list, to the suffix of the list, to the suffix of the suffix and so on. Interestingly,
it preserves the \emph{shape} of the list as it turns a list of $n$ elements into another list
of $n$ elements.

@@

Now, consider a simple data-flow language with single-variable contexts, variables,
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$
has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
the language models data-flow and values are lists (of past values) is a matter of semantics,
which is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (extract the variable value); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).
For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an
expression that expects context with variable $x$ and returns a stream of values before the
previous one. The semantics of the term expresses this behaviour:
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed
earlier.

@@

We return to this topic when discussing lambda abstraction in Section~\ref{sec:applications-structure-lam}
and semantics of flat coeffect systems in Section~\ref{sec:semantics-theory}.

@@

\paragraph{Meta-language interpretation.} To demonstrate the approach that employs comonads
as part of a meta-language, we look at an example inspired by the work of Pfenning et al.
\cite{logic-modal-reconstruction,logic-cmtt}. We do not attempt to provide a precise overview of
their work. The main purpose of the following discussion is to provide a different intuition behind
comonads, and to present an example of a language that includes comonad as a type constructor,
together with language primitives corresponding to comonadic operations\footnote{In fact,
Pfenning et al. never mention comonads explicitly. This is done in later work by Gabbay et al.
\cite{logic-cmtt-semantics},  but the connection between the language and comonads
is not as direct as in case of monadic or comonadic semantics covered in the previous section.}.

@@

In the coeffect system for tracking resource usage outlined earlier, we associated additional
contextual information (set of available resources) with the variable context of the typing
judgement: $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$. In other words, our work focuses on
what is happening on the left hand side of $\vdash$.
